---
title: "AE03 - Exploring gerrymandring"
author: "Aliyah Cardenas"
date: "2026-09-10"
format: html
execute: 
  warning: false
  message: false
---

## Getting started

Load the `tidyverse`, `usdata`, and `sjPlot` packages, and get the `gerrymander` data out of the `usdata` package. View the data using `glimpse`.

```{r}
library(tidyverse)
library(usdata)
library(sjPlot)
library (gtsummary)

gerrymander <- usdata::gerrymander
```

## 1. Congressional districts per state

Which state has the most congressional districts? The least? How many congressional districts are there in this state? Use functions such as `table`, `tbl_summary` or `count`.

```{r}
gerrymander %>%
  count(state) %>%
  arrange(desc(n))
  
```
California has the most congressional districts (53). Multiple states (i.e. WY, VT, SD, ND) have the lowest amount at 1.

## 2. Districts at the tails

Make side-by-side box plots of percent of vote received by Trump in the 2016 Presidential Election by prevalence of gerrymandering. Identify any Congressional Districts that are potential outliers — points that fall well outside the whiskers of their box, meaning that district's Trump vote share was unusually high or low compared to other districts with the same gerrymandering prevalence. What information can you share about them?

```{r}
ggplot(gerrymander, aes(x = gerry, y = trump16))+
  geom_boxplot()

gerrymander %>%
  filter(gerry == "low", trump16 < 10) %>%
  select(state, district, trump16, gerry)
```

The outliers that fall below the lower whisker of the low gerrymandering box are districts NY-13 and NY-15. Trump only got 5.4% of the vote from NY-13 and 4.9% in NY-15 which is much lower compared to other low districts.

## 3. Gerrymandering and flipping

This relationship can be looked at from two directions — does gerrymandering predict a 2018 flip, and does a 2018 flip predict gerrymandering? Support **both** with a visualization and summary statistics. The plot answers "what does the mix look like within each group?"; the table gives the exact proportions the plot is showing. Write 1-2 bullet points about what you noticed from each plot. Use specific numbers in your answer.

**Direction 1:** Is a Congressional District more likely to have high prevalence of gerrymandering if a Democrat was able to flip the seat in the 2018 election? To compare the breakdown of gerrymandering prevalence within each flipped group, make sure `flip18` is on the x axis and that your percents add up to 100% within each category.

```{r}
ggplot(gerrymander, aes(x = flip18, fill = gerry)) +
  geom_bar(position = "fill")

gerrymander %>%
  count(flip18, gerry) %>%
  group_by(flip18) %>%
  mutate(prop = n / sum(n))

```
- In 2018, districts where a democrat was able to flip the seat had a lower proportion of high gerrymandering prevalence at 13.2%, compared to the 25.0% prevalence in districts that didn't flip.

**Direction 2:** Is a Congressional District more likely to be flipped to a Democratic seat if it has high prevalence of gerrymandering, or low prevalence? Same two variables, just flipped — so now `gerry` is on the x-axis. Again, make sure your percents add up to 100% within each category.

```{r}

ggplot(gerrymander, aes(x = gerry, fill = factor(flip18)))+
  geom_bar(position = "fill")

gerrymander %>%
  count(gerry, flip18) %>%
  group_by(gerry) %>%
  mutate(prop = n / sum(n))


```
- Districts are more likely to be flipped to a Democratic seat (at 12.9%) if they have a low gerrymandering prevalence compared to districts with mid (at 9.3%) or high prevalence (4.9%)

## 4. Gerrymandering and presidential margin

Using `mutate`, create a new variable, `margin16`, measuring how many percentage points Clinton won or lost by in each district (`clinton16 - trump16`). Does gerrymandering prevalence relate to how lopsided the 2016 presidential race was in that district? Report the min, mean, and max margin by `gerry` level, then visualize the full distribution with a violin + boxplot combo. Describe the shape, center, and spread of each group.

```{r}
gerrymander %>%
  mutate(margin16 = clinton16 - trump16) %>%
  group_by(gerry) %>%
  summarise(
    min_margin16 = min(margin16),
    mean_margin16 = mean(margin16),
    max_margin = max(margin16))

gerrymander%>%
  mutate(margin16 = clinton16 - trump16) %>%
ggplot(aes(x = gerry, y = margin16))+
  geom_violin() + geom_boxplot(width = .4)

```
- Low: This gerrymandering level has a bimodal distribution where the larger concentration lays at negative margins at around ~-10, and the other is at a large positive margin at around ~70. For the center, it has a mean margin of 6.3 points. It  has a wide  spread with a range from -54.9 to 88.9.
- Mid: This looks unimodal (roughly) with the largest distribution laying at around ~-15. The mean margin is 3.3 and it has a a wide spread from -63 to 82.8.
- High: This distribution is bimodal with the slightly larger concentration laying just around the same margin as the mid gerrymandering group. Its mean margin is 1.3 points and it ranges from -63 to 78.2.

In districts with low gerrymandering prevalence the margins were closest to 0, but as we move up the groups, it narrows near 0 and gets wider further away from it. This could imply a relationship where higher gerrymandering correlates to a more lopsided race
